A weighted sampling algorithm for the design of RNA sequences with targeted secondary structure and nucleotides distribution

### 2013-03-10 ISMB/main_ISMB.tex
fix availability
Typo in equations + Formatting
Protected biblio

\title{A weighted sampling algorithm for the design of RNA sequences with targeted secondary structure and nucleotides distribution}
\author{Vladimir Reinharz$^1$, Yann Ponty$^{2,*}$, J\'er\^{o}me Waldisp\"{u}hl$^{1,*}$\footnote{to whom correspondence should be addressed}}
\address{$^1$ School of Computer Science, McGill University, Montreal, Canada\\ $^2$ Laboratoire d'informatique, \'Ecole Polytechnique, Palaiseau, France.}

@@

\section{Results:}
In this paper, we introduce \ourprog, a novel algorithm to design RNA sequences folding into target secondary structures with a predefined nucleotide distribution.  \ourprog uses a global sampling approach and weighted sampling techniques. We show that our approach is fast (i.e. running time comparable or better than local search methods), seed-less (we remove the bias of the seed in local search heuristics), and successfully generates high-quality sequences (i.e. thermodynamically stable) for any \GCContent. To complete this study, we develop an hybrid method combining our global sampling approach with local search strategies. Remarkably, our glocal methodology overcomes both local and global approaches  for sampling sequences with a specific GC content and target structure.
\section{Availability:} 
Our method is freely available at WEBSITE INCARNATION
\section{Contact:} \href{jeromew@cs.mcgill.ca}{jeromew@cs.mcgill.ca}, \href{yann.ponty@lix.polytechnique.fr}{yann.ponty@lix.polytechnique.fr}\\
\noindent
\textbf{Key words:} RNA, secondary structure, design, weighted sampling, GC-content.
\end{abstract}

>>>>

\title{A weighted sampling algorithm for the design of RNA sequences with targeted secondary structure and nucleotides distribution}
\author{Vladimir Reinharz$^1$, Yann Ponty$^{2,*}$, J\'er\^{o}me Waldisp\"{u}hl$^{1}$\footnote{to whom correspondence should be addressed}}
\address{$^1$ School of Computer Science, McGill University, Montreal, Canada\\ $^2$ Laboratoire d'informatique, \'Ecole Polytechnique, Palaiseau, France.}

@@

\section{Results:}
In this paper, we introduce \ourprog, a novel algorithm to design RNA sequences folding into target secondary structures with a predefined nucleotide distribution.  \ourprog uses a global sampling approach and weighted sampling techniques. We show that our approach is fast (i.e. running time comparable or better than local search methods), seed-less (we remove the bias of the seed in local search heuristics), and successfully generates high-quality sequences (i.e. thermodynamically stable) for any \GCContent. To complete this study, we develop an hybrid method combining our global sampling approach with local search strategies. Remarkably, our glocal methodology overcomes both local and global approaches  for sampling sequences with a specific GC content and target structure.
\section{Availability:} 
\ourprog is available at \href{http:// csb.cs.mcgill.ca/incarnation/}{csb.cs.mcgill.ca/incarnation/}
\section{Contact:} \href{jeromew@cs.mcgill.ca}{jeromew@cs.mcgill.ca}, \href{yann.ponty@lix.polytechnique.fr}{yann.ponty@lix.polytechnique.fr}\\
\noindent
\textbf{Key words:} RNA, secondary structure, design, weighted sampling, GC-content.
\end{abstract}

### 2013-03-10 ISMB/methods_ISMB.tex
Protected biblio

The recursion must consider different cases, depending on whether the first position in $\Struct$ is:
\begin{itemize}
\item {\bf Case 1.} Unpaired ($\Struct = \ub \Struct'$):
\begin{align}
	\Z{\BoolTrue,\ub \Struct'}{a,b} = \Z{\BoolFalse,\ub \Struct'}{a,b}  &:=
      \sum_{a'\in \B}  
      x^{\gc(a')}
      \cdot\Z{\BoolFalse,\Struct'}{a',b}; 
\label{eq:Z_unpaired}
\end{align}
\item {\bf Case 2.} Paired to last position ($\Struct = \op \Struct' \cp$), stacking onto a pre-existing closing pair ($\N=\BoolTrue$):
\begin{align}
	\Z{\BoolTrue,\op \Struct' \cp}{a,b} &:=
      \sum_{a',b'\in \B^2}
			 x^{\gc(a'.b')}
			 \cdot e^{\frac{-\ES^{\beta}_{ab \to a'b'}}{RT}}
			 \cdot \Z{\BoolTrue,\Struct'}{a',b'};
\label{eq:Z_stack}
\end{align}
\item {\bf Case 3.} Paired to some position ($\Struct = \op \Struct' \cp \Struct''$), but not stacking ($\N=\BoolFalse$ or $\Struct''\neq \varepsilon$):
\begin{align}
	\Z{\N,\op \Struct' \cp \Struct''}{a,b} &:=
			 \displaystyle
      \sum_{a',b'\in \B^2}
      x^{\gc(a'.b')}
			\cdot e^{\frac{-\ES^{\beta}_{\varnothing\to a'b'}}{RT}}
      \cdot\Z{\BoolTrue,\Struct'}{a',b'}
      \cdot\Z{\BoolFalse,\Struct''}{b',b}.
\label{eq:Z_rec}
\end{align}
\end{itemize}

@@

\begin{algorithm}[t]
\DontPrintSemicolon
	\SetAlgoLined
\SetKwFunction{Backtrack}{SB}
\SetKwFunction{Random}{Random}
	rand $\leftarrow$ \Random$\left(\Z{\N,\Struct}{a,b}\right)$\tcp*[r]{Draw random number in $[0,\Z{\N,\Struct}{a,b}[$}
 \Switch{}{
   \lCase(\tcp*[f]{Empty structure $\Rightarrow$ empty sequence}){$\Struct=\varepsilon$}{\Return{$\varepsilon$}}
   \Case(\tcp*[f]{First position is unpaired}){$\Struct=\ub\, \Struct'$}{		
   \For{$a'\in\B$}{
			rand $\leftarrow$ rand $- x^{\gc(a')}\cdot \Z{\BoolFalse,\Struct'}{a',b}$\;
			\lIf{$\text{\rm rand}<0$}\Return{$a'.\Backtrack(a',b,\BoolFalse,\Struct')$}\;
  		}
   }
   \Case(\tcp*[f]{Extremities involved in stacking base pair}){$\Struct=\text{\rm\op}\, \Struct' \,\text{\rm\cp}$  {\rm \bf and} $\N=\BoolTrue$}
   {		
			\For{$(a',b')\in\B\times\B$}
    {
				rand $\leftarrow$ rand $-
			 	x^{\gc(a'.b')}
			 	\cdot e^{{-\ES^{\beta}_{ab \to a'b'}}/{RT}}
			 	\cdot \Z{\BoolTrue,\Struct'}{a',b'}	$\;
				\lIf{$\text{\rm rand}<0$}{
					\Return{$a'.\Backtrack(a',b',\BoolTrue,\Struct'). b'$}\;		
				}
			}
		}
   \Other(\tcp*[f]{First position is paired without a stacking pair})
   {
				\tcp{$\Struct=\text{\rm\op}\, \Struct' \,\text{\rm\cp}\,\Struct''$}
				\For{$(a',b')\in\B\times\B$}{
					rand $\leftarrow$ rand $-	
       x^{\gc(a'.b')}
					\cdot e^{\frac{-\ES^{\beta}_{\varnothing\to a'b'}}{RT}}
  	    	\cdot\Z{\BoolFalse,\Struct'}{a',b'}
	    	  \cdot\Z{\BoolTrue,\Struct''}{b',b}$\;	
 					\lIf{{\rm rand} $<0$}{
						\Return{$a'
						.\Backtrack\left(a',b',\BoolFalse,\Struct'\right)
						.b'
						.\Backtrack\left(b',b,\BoolTrue,\Struct''\right)
						$}\;	
					}
				}
   }
 }
\caption{\protect\Backtrack$\left(a,b,\N,\Struct\right)$\label{alg:back}}
\end{algorithm}
\subsubsection{Stochastic backtrack}
Once the pseudo-partition functions have been computed and memorized, a stochastic backtrack starts from the target structure $\Target$ with any exterior bases $[a,b]$ and no nesting base-pair, corresponding to a call \Backtrack$\left(\varnothing,\varnothing,\BoolFalse,\Target\right)$ to Algorithm~\ref{alg:back}.  At each step, a suitable assignment for one or several positions is chosen, using probabilities derived from the precomputation, as illustrated by Figure~\ref{fig:stochastic}. One or several recursive calls over the appropriate substructures are then performed. On each recursive call, the algorithm assigns at least one nucleotide to a -- previously unassigned -- position. Moreover, the number of executions of each loops is bounded by a constant. Consequently, the complexity of Algorithm~\ref{alg:back} is in $\Theta(n)$ time and space.

\subsubsection{Self-adaptive sampling strategy and overall complexity}

@@

Regarding the overall complexity claim for our approach, it was previously established~\cite{Waldispuhl2011} that, for each value of $x$, there exists constants $\mu_x$ and $\sigma_x$ such that the distribution of \GCContent asymptotically converges towards a normal law having expectation in $\mu_x\cdot n\cdot(1+o(1))$ and standard deviation in $\sigma_x\cdot\sqrt{n}\cdot(1+o(1))$.
Furthermore, the concentration of the distribution, asserted by its limited standard deviation, therefore the expected number of attempts required to generate a valid sequence when $\kappa=0$ (resp. $\kappa\in\Omega(1/\sqrt n)$) grows like $\Theta(\sqrt{n})$ (resp. $\Theta(1)$, i.e. a constant), leading to the announced complexities. Formally, since a suitable weight $x$ must be recomputed for each targeted structure and \GCContent, then the number $M$ of iterations required for the converge can be accounted for explicitly, leading to time complexities in $\Theta((M+\sqrt{n})\cdot k\cdot n)$ (if $\kappa=0 \Rightarrow$ no tolerance) and $\Theta(M\cdot k\cdot n)$ (if $\kappa>0$).

>>>>

The general recursion scheme specializes, depending on the first position in $\Struct$.\\
{\bf Case 1.} First position is unpaired ($\Struct = \ub \Struct'$):
\begin{equation}
	\Z{\BoolTrue,\ub \Struct'}{a,b} = \Z{\BoolFalse,\ub \Struct'}{a,b}  :=
      \sum_{a'\in \B}  
      x^{\gc(a')}
      \cdot\Z{\BoolFalse,\Struct'}{a',b}; 
\label{eq:Z_unpaired}
\end{equation}
{\bf Case 2.} First position is paired with last position ($\Struct = \op \Struct' \cp$), stacking onto a pre-existing exterior pair ($\N=\BoolTrue$):
\begin{equation}
	\Z{\BoolTrue,\op \Struct' \cp}{a,b} :=
      \sum_{a',b'\in \B^2}
			 x^{\gc(a'.b')}
			 \cdot e^{\frac{-\ES^{\beta}_{ab \to a'b'}}{RT}}
			 \cdot \Z{\BoolTrue,\Struct'}{a',b'};
\label{eq:Z_stack}
\end{equation}
{\bf Case 3.} First position is involved in a base-pair ($\Struct = \op \Struct' \cp \Struct''$), which is not stacking onto an exterior base-pair ($\N=\BoolFalse$ or $\Struct''\neq \varepsilon$):
\begin{equation}
	\Z{\N,\op \Struct' \cp \Struct''}{a,b} :=
			 \displaystyle
      \sum_{a',b'\in \B^2}
      x^{\gc(a'.b')}
			\cdot e^{\frac{-\ES^{\beta}_{\varnothing\to a'b'}}{RT}}
      \cdot\Z{\BoolTrue,\Struct'}{a',b'}
      \cdot\Z{\BoolFalse,\Struct''}{b',b}.
\label{eq:Z_rec}
\end{equation}

@@

\begin{algorithm}[t]
\DontPrintSemicolon
	\SetAlgoLined
\SetKwFunction{Backtrack}{SB}
\SetKwFunction{Random}{Random}
\newcommand{\rand}{{r}}
	$\rand \leftarrow $\Random$\left(\Z{\N,\Struct}{a,b}\right)$\tcp*[r]{Random real in $[0,\Z{\N,\Struct}{a,b}[$}
 \Switch{}{
   \lCase(\tcp*[f]{Empty structure}){$\Struct=\varepsilon$}{\Return{$\varepsilon$}}
   \Case(\tcp*[f]{First position is unpaired}){$\Struct=\ub\, \Struct'$}{		
   \For{$a'\in\B$}{
			$\rand \leftarrow \rand - x^{\gc(a')}\cdot \Z{\BoolFalse,\Struct'}{a',b}$\;
			\lIf{$\rand<0$}\Return{$a'.\Backtrack(a',b,\BoolFalse,\Struct')$}\;
  		}
   }
   \Case(\tcp*[f]{Extremities are involved in stacking base pair}){$\Struct=\text{\rm\op}\, \Struct' \,\text{\rm\cp}$  {\rm \bf and} $\N=\BoolTrue$}
   {		
			\For{$(a',b')\in\B\times\B$}
    {
				$\rand \leftarrow \rand -
			 	x^{\gc(a'.b')}
			 	\cdot e^{{-\ES^{\beta}_{ab \to a'b'}}/{RT}}
			 	\cdot \Z{\BoolTrue,\Struct'}{a',b'}	$\;
				\lIf{$\rand<0$}{
					\Return{$a'.\Backtrack(a',b',\BoolTrue,\Struct'). b'$}\;		
				}
			}
		}
   \Other(\tcp*[f]{First position is paired without a stacking pair})
   {
				\tcp{$\Struct=\text{\rm\op}\, \Struct' \,\text{\rm\cp}\,\Struct''$}
				\For{$(a',b')\in\B\times\B$}{
					$\rand \leftarrow \rand -	
       x^{\gc(a'.b')}
					\cdot e^{\frac{-\ES^{\beta}_{\varnothing\to a'b'}}{RT}}
  	    	\cdot\Z{\BoolFalse,\Struct'}{a',b'}
	    	  \cdot\Z{\BoolTrue,\Struct''}{b',b}$\;	
 					\lIf{\rand $<0$}{
						\Return{$a'
						.\Backtrack\left(a',b',\BoolTrue,\Struct'\right)
						.b'
						.\Backtrack\left(b',b,\BoolFalse,\Struct''\right)
						$}\;	
					}
				}
   }
 }
\caption{\protect\Backtrack$\left(a,b,\N,\Struct\right)$\label{alg:back}}
\end{algorithm}
\subsubsection{Stochastic backtrack}
Once the pseudo-partition functions have been computed and memorized, a stochastic backtrack starts from the target structure $\Target$ with any exterior bases $[a,b]$ and no nesting base-pair, corresponding to a call \Backtrack$\left(\varnothing,\varnothing,\BoolFalse,\Target\right)$ to Algorithm~\ref{alg:back}.  At each step, a suitable assignment for one or several positions is chosen, using probabilities derived from the precomputation, as illustrated by Figure~\ref{fig:stochastic}. One or several recursive calls over the appropriate substructures are then performed. On each recursive call, the algorithm assigns at least one nucleotide to a -- previously unassigned -- position. Moreover, the number of executions of each loops is bounded by a constant. Consequently, the complexity of Algorithm~\ref{alg:back} is in $\Theta(n)$ time and space.

\subsubsection{Self-adaptive sampling strategy}

@@

\subsubsection{Overall complexity}

Regarding the overall complexity claim for our approach, it was previously established~\cite{Waldispuhl2011} that, for each value of $x$, there exists constants $\mu_x$ and $\sigma_x$ such that the distribution of \GCContent asymptotically converges towards a normal law having expectation in $\mu_x\cdot n\cdot(1+o(1))$ and standard deviation in $\sigma_x\cdot\sqrt{n}\cdot(1+o(1))$.
Furthermore, the concentration of the distribution, asserted by its limited standard deviation, therefore the expected number of attempts required to generate a valid sequence when $\kappa=0$ (resp. $\kappa\in\Omega(1/\sqrt n)$) grows like $\Theta(\sqrt{n})$ (resp. $\Theta(1)$, i.e. a constant), leading to the announced complexities. Formally, since a suitable weight $x$ must be recomputed for each targeted structure and \GCContent, then the number $M$ of iterations required for the converge can be accounted for explicitly, leading to time complexities in $\Theta((M+\sqrt{n})\cdot k\cdot n)$ (if $\kappa=0$, i.e. without any tolerance) and $\Theta(M\cdot k\cdot n)$ (if $\kappa>0$).
